%*******************************************************
% Acknowledgments
%*******************************************************

\cleardoublepage

\phantomsection
\addcontentsline{toc}{chapter}{\tocEntry{Acknowledgments}}

\vspace*{-1.4cm}
\begin{addmargin}{-1.0cm}

\begin{flushright}
    \textsl{\onehalfspacing
    The meeting of two personalities \\
    is like the contact of two chemical substances:\\
    if there is any reaction, both are transformed.}\\
    --- Carl Gustav Jung
\end{flushright}

\bigskip

\begingroup
\setstretch{1.3}

% \let\clearpage\relax
\let\cleardoublepage\relax
% \let\cleardoublepage\relax

\vspace*{-0.7cm}
\chapter*{Acknowledgments}
\vspace*{-0.6cm}

Ups and downs. Research, much like life, is a process.
We often expect a smooth and straightforward path, only to find it filled with unexpected hurdles, surprising twists, and -- occasionally -- those truly remarkable moments at the summit.

As I embarked on this journey, I was already immersed in the vibrant world of research at the Dynamic Interaction Control laboratory within the prestigious walls of {\small IIT}, guided by Francesco Nori.
To him, I extend my heartfelt gratitude for giving me the opportunity to be part of this beautiful, unique, and enriching community, which has been my home for the past seven years.
The memories forged in Genova -- professional and personal -- will forever shape the narrative of my life.

My deepest appreciation goes to my supervisor, Daniele Pucci, head of the Artificial Mechanical Intelligence laboratory.
Daniele, you have been not only a supervisor but also a colleague, a friend, and, as time passed, a mentor.
I am profoundly thankful for the freedom you granted me to explore uncharted territories without the weight of mere numerical publications.
Your support, even for the craziest ideas -- some good, some not so much :) -- and your guidance in navigating the most intricate of situations have been invaluable.
I must also express my gratitude to Professor Angelo Cangelosi for his supervision from the Manchester side and his constant availability to clarify doubts and provide insights.
For the necessary funding that sustained my research project, I owe an immense debt of gratitude to Giorgio Metta.
His support was pivotal in bringing my research to fruition.
I cannot thank Professor Bruno Adorno and Professor Michael Mistry enough for their roles as examiners during my viva.
Your feedback has been instrumental in achieving the final quality of this thesis.
I'm grateful for your friendly approach in the rigorous context of the defense, you managed to transform it into a meaningful and enjoyable discussion among peers.
I couldn't have asked for a more supportive and constructive examination.

Listing all the individuals who have played pivotal roles in my journey is an impossible task.
In AMI, we've grown too vast for individual mentions, but it's essential to convey how profoundly meaningful your impact has been.
Extending this sentiment to the broader IIT community, I've had the privilege of learning from brilliant scientists and talented engineers.
I've absorbed so much knowledge from our interactions, and I hope to have given back even a fraction of what I received.
Navigating this world of research together has been challenging, but it's a journey that's enriched us all.
I'm confident that each of you is destined for great success, and our future accomplishments will be a direct result of this collaborative journey.
Thank you, friends.

Zooming out for a broader perspective, the contributions highlighted in this thesis owe much to the continuous efforts of the entire robotics and open-source communities. \linebreak
These communities provide essential tools and shared knowledge.
Research is done by tiny bits from individuals, but these bits would have no meaning if not built into a broader, well-established context.
I hope that with this work, I succeeded in contributing my grain to this vast sandy beach of knowledge.
GitHub, in particular, has become a cozy home where to collaborate, strengthen our efforts, make connections, and foster growth.
I wish other domains could embrace such a positive spirit.

I hold a strong belief that I wouldn't have been able to weather with the countless lows that characterize most doctoral journeys and maintain my mental well-being without the relief that sports and the great outdoors provided.
When I felt out of place, amidst the rigors of day-to-day life, spending time in nature became essential to ground me down again.
Out there, I feel more alive than words can convey; it's as if I'm deeply rooted to my origins, and these roots don't thrive within the confines of concrete. \linebreak
I'm sincerely thankful for the tranquility of the mountains, which reflects the inner peace I aim to experience daily and represents the place where I feel most at home.
Those who can't understand this connection may find it challenging to comprehend who I am. \linebreak
In the winter, I find comfort in the Alps, rejuvenating through the simple act of skiing with its corresponding rush of adrenaline.
As warmer seasons approach, I embark on countless hikes, exploring the stunning landscapes of Liguria or my beloved Friuli mountains back home.
Thanks to Italy, with its incredible diversity.
Food, nature, culture, people, weather - life outside the office is truly where it's at.
There's no better place in the world where all of these elements come together so harmoniously.

Throughout the year, CrossFit has always been a constant presence, providing a regfuge for both mental and physical renewal, and offering short-term goals that kept me on track.
Amidst the challenges, whether the demands of academia or the disruptions of a global pandemic, the comfort drawn from these activities has been my saving grace.
To Canteri CrossFit, you hold a special place in my heart.
I found teammates, passionate coaches, good friends, and a new family within your four walls.
There were times when I walked into the box feeling well over my head, I admit, but all doubts and worries dissipated as I walked out, stronger in every way.
You've been my lifeline, safeguarding my mental health during these years, and I cannot express enough gratitude for your unwavering support.

And last but certainly not least, I want to conclude by extending a heartfelt thank you to my family: my mom Silvana, my sister Erica, and my brother Giulio.
Your consistent support and unwavering acceptance of every choice I've made throughout my life, without judgment, has been a source of immeasurable strength.
Dad, even though you couldn't witness this journey firsthand, I always felt my back covered by your usual silent approval. I love you all.

\end{addmargin}

\vfill
\endgroup
